Inefficient Concessions and Mediation

### 2023-08-20 draft.tex
rename working document draft_narrow back to draft

\begin{abstract}
When two parties are engaged in conflict and each distrusts the other's ability to pursue peace, concessions can be used to indicate an interest in making peace. However, when negotiating parties are concerned that concessions could be used against them in the future, a lack of trust can prevent optimal concessions from being made. It can also reduce the possibility of peace. We use mechanism design to explore ways in which a third party mediator can act as a guarantor that promised concessions will be delivered, thereby reducing inefficiencies and increasing the potential for peace. In this process, we open up a new rationale for mediation: to increase the inefficiency of signaling in a preliminary round of negotiations. 
\end{abstract}

@@

In the course of a military conflict, settlement negotiations often occur simultaneously with continued fighting. If settlement can be achieved, lives are saved and destruction averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (e.g., Bercovitch 2000; Greig 2005a; Beardsley et al. 2006; Duursma 2020).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that combatants can offer. In our context, the efficiency of a concession is determined by how much of the material value of the concession is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

We introduce a two-sided signaling model in which concessions can provide future value to the recipient and/or serve as a signal that provides the recipient with information about the type of player the conceding state is. Using a mechanism design framework, we demonstrate that a mediator can facilitate more efficient concessions by removing uncertainty about the ability of the parties to commit to peace. Perhaps more importantly, the mediator can also sometimes facilitate peace when the negotiating parties cannot achieve it through bilateral interaction.

Two archetypal situations illustrate the significance and difficulty of efficient concession-making in war settlement.  The first is the case of disarmament and the second is the concession of control over strategically valuable territory.

Barbara Walter identified disarmament as the ``critical barrier to civil war settlement'' (1997). Particularly in conflicts between a rebel group and a central government, the best-achievable negotiated settlement often involves policy concessions by the government in exchange for disarmament and recognition of government authority by the rebels. Disarmament by rebels is an efficient concession if fully-operational weapons are turned over to the central government for its own use, while disarmament accompanied by decommissioning is inefficient because the value of the arms is destroyed. Disarming and turning over fully functional weapons is an incredibly difficult concession for rebels to make: once they turn over their arms, they lose future bargaining leverage and have no guarantee that their own weapons won't later be used against them. If the rebels can instead decommission those weapons, an important stumbling block to peace is removed.\footnote{Other stumbling blocks remain, including the fact that government retains its weapons. Thus, credible commitment on the part of the government remains difficult, even if decommissioning occurs.} Peace may therefore be possible with inefficient concessions where it is impossible with efficient concessions.

Control of strategically valuable territory can pose similar challenges in both civil and interstate wars. Even if transferring control of strategically valuable territory creates the best chance for peace, its transfer often involves a substantial risk that the territory will provide military advantage in a future conflict.

Following the 1992 war between Armenia and Azerbaijan over the disputed region of Ngorno-Karabakh, Armenia retained control of a significant amount of Azeribaijani territory that neither the Armenian government nor the Karabakh government viewed to be part of its homeland. Azerbaijan, however, valued control of the territory greatly as the home of Azeri residents and wished to govern the territory. Thus, it might have been efficient in the mid-1990s for the Armenian/Karabahk side to cede control of that territory to Azerbaijan as part of a deal in which Azerbaijan recognized Karabakh's independence.  However, that territory was strategically valuable, including territory on both sides of the Lachin Corridor, which connects Ngorno-Karabakh and Armenia. Azerbaijani control of the territory surrounding the corridor would empower Azerbaijan to seize military advantage in the future and cut off land travel between Ngorno-Karabakh and Armenia.  The territory was not conceded and the conflict persisted in costly stalemate for almost twenty years until war broke out anew in 2020.

Walter (1997) observes empirically that, due to the difficulty of making efficient disarmament concessions, civil wars tend to end on the battlefield, rather than through negotiated settlement. Interstate wars over territory are similarly prone to violent, rather than negotiated, resolution (Hensel 1996). Can mediation offer a path to settlement in these challenging contexts?

This paper revises our understanding of the reasons why efficient concessions are sometimes impossible to make, why inefficient concessions can sometimes be beneficial, and how mediators can both help increase the efficiency of concessions and make peace. Weapons or strategically valuable territory will not be conceded if the conceding side fears that territory or weapons may be used in future attacks against them. One way to address this concern is to destroy the future value of the concession---e.g. decommission weapons or attempt to reduce the strategic value of the territory.  Another option, and the one on which we focus in this paper, is third-party mediation. We use this model to demonstrate that a third-party mediator can facilitate negotiated settlement by increasing trust and enabling combatants to make more efficient concessions, even when there is a risk those concessions might increase the bargaining power of their opponent in the future.

While an integral part of the literature on conflict, concessions have been only partially addressed by theoretical research. The broader literature that attempts to model war has included concessions as part of bargaining, but has not explicitly considered the role of concessions in achieving peace. Schelling's \emph{Arms and Influence} (1966) modeled war with a game theoretic framework, and Powell (1999) has also defended the use of modeling in understanding war. Fearon (1996), Powell (1996a and 1996b, 2006) and others consider war to be a bargaining failure, caused by commitment problems and asymmetric information. When the assets that are the object of the bargaining confer future power, a lack of credible commitment, or trust, can create a holdup problem that prevents the optimal level of cooperation (Fearon 1996).

In these models war can be averted by splitting the pie (i.e. the disputed territory) in a way acceptable to both parties. Concessions are modeled as offering part of the pie to the other party, and asymmetric information is about each party's \textit{resolve} to fight and/or military capabilities (Arena 2013). In this context, an example of costly signaling would be military preparations for war (Slantchev 2011).

We will approach the commitment problem through this same general understanding, but we model a party's ability to cooperate as private information. A high type is able to cooperate; a low type has incentives not to cooperate. These types map onto the commitment problem, with a high type one who is able to commit in certain scenarios. A low type, though perhaps willing, is unable to do so. A low type may posture like the high type, but lacks the political muscle or will to follow through on a peace agreement. Democratic leaders may more often be high types because they are less likely to face coups or other hostilities in reaction to short term decisions. This is not always the case, however. Mahmous Abbas may be a low type, not because he does not want to make peace, but because he cannot credibly act on behalf of the entire Palestinian population.

In conflict resolution, concessions may act as signals or as an end in themselves -- or as both simultaneously. One contribution of this paper is that it allows concessions to have future material value, signaling value, and both. We will show that when concessions have both signaling and future material value, some of the material value of the concessions may be purposefully destroyed and thus inefficiencies can emerge.

In analyzing both the signaling and material value of concessions, an extensive literature on gift giving is particularly helpful. Like concessions, costly gifts have been shown to have a valuable role in relationship building. And the gift giving literature, more so than the concessions literature, focuses specifically on the issue of efficiency. Inefficiency is defined for the purposes of this paper as giving a gift (i.e. a concession) that is more costly to the giver than it is valuable to the recipient.

Often, inefficient gifts occur in the formative stages of relationships or in immature relationships (Camerer 1988). Gifts given in established relationships, such as wedding presents or spousal gifts, are more likely to be cash gifts or efficient gifts that are specifically requested or selected from a registry. Since inefficient gift giving is more likely involved in nascent relationships and partnerships, it is natural to suppose that inefficient gifts play a signaling role in addition to the material value of transferring the gift's inherent worth.

In efforts to explain the existence of gift giving,  Camerer (1988) and Van de Ven (2000) give several anthropological explanations for inefficiency. Amongst these explanations are altruism, social mores, and egoism. But the most relevant explanation for the field of international relations is that gift giving is strategic. Camerer uses mechanism design to model gifts as costly signals; inefficiency is useful in pre-play communication in order to form relationships between like types. This strategic giving of gifts in societies looks very much like the strategic giving of concessions to remedy conflict situations. When modeled game-theoretically, these environments look even more similar.

@@

Camerer shows that in this general formulation, a costly signaling model will always yield efficient gifts.\footnote{In an environment with repeated opportunities for concessions Watson (1999) is able to show that a strong relationship can be reached by starting with small ``gifts'' and increasing their size as the relationship becomes more committed. See Van de Ven (2000) for an overview of this literature.} In order to explain inefficiency, Camerer adds an additional pre-play period where players must pay a cost in order to enter the game and send and receive gifts. If types separate in the pre-play period with low types not being willing to pay, high types are saved the cost of sending a signal to low types in the main game.  High types would, in some parameterizations, be given incentives to give inefficient equilibrium gifts in the main game in order to lower the low types' expected payoff of playing to below the pay-to-play fee. Such an action can keep out the low types and this can raise the expected payoff for a high type by saving on the gifts given.

Prendergast and Stole (2001) provide a second explanation for inefficient gifts. The authors explain inefficient gift giving by modeling the utility of matching. That is, if players are of a like mindset they get an extra benefit so they are willing to take the risk of giving the wrong gift.

We propose a different conceptual framework to explain this real-life phenomenon that is rooted in a realistic international relations puzzle. The basic formulation of our mechanism-design approach is an expansion of Camerer's (1988) analysis. The model can achieve the basic `inefficient gifts' results of Camerer (1988) and Prendergast and Stole (2001) but does so using a very different explanation for inefficiency. The basic model is a costly signaling model, at its root similar to Spence (1973) and subsequent papers.\footnote{See Connelly et al. (2011) for a review of the costly signaling literature.} In essence, the model shows that if concessions can be later used against the party that gives them, then it can be in a negotiating party's best interest to give inefficient concessions. While our model is tailored to an international relations setting, the explanation for inefficient giving of concessions may be more broadly applicable.

@@

It is essential to focus more precisely on the specific actions and environments that make use of particular types of mediation, which is how we approach the question. This paper models an environment where the primary issue is trust, not uncertainty about an opponent's capabilities, and where a mediator is employed who has manipulative abilities. Mediation rarely occurs in the absence of violent conflict (Bercovitch 1996; Beardsley 2006), and parties are more likely to seek mediation when they are entrenched in costly conflict (Bercovitch and Jackson 2001; Greig 2005b; Greig and Diehl 2006; Terris and Maoz 2005; Svensson 2008). We thus model parties to a conflict who can benefit from seeking mediation in order to overcome mistrust.\footnote{Alternatively, parties may seek a mediator to justify difficult concessions and avoid angering domestic constituencies (Allee and Huth 2006, Beardsley 2010; Beardsley and Lo 2014).}

These empirical results have until recently lacked strong theoretical foundations. Mediation is not uncommon, but we are only beginning to understand why actors use it. Recently, mediation has begun to be modeled using mechanism design for settings where militarization or the resolve to fight are private information (Bester and Warneryd 2006; Fey and Ramsay 2009, 2010 and 2011; Horner et. al. 2010; Meirowitz et al. 2012, 2015). This paper expands the theoretical literature by modeling the role of mediation when information is asymmetric about parties' abilities to commit to peace. Here, a mediator can remove the incentive to make concessions that are inefficient.

@@

Beginning in period 1, countries engage in an infinitely repeated Prisoner's Dilemma once per period with stage game payoffs in Table 1 below.\footnote{For a game-theoretic analysis of the repeated Prisoner's Dilemma with uncertainty over discount factors but no signaling game, see Maor and Solan (2015). They focus on the players' belief structures, while our initial signaling stage simplifies the analysis of beliefs.} While not a good representation of all-out war or even a zero-sum short term war, the Prisoner's Dilemma is appropriate to represent a limited, continued conflict. In cases appropriately modeled by a Prisoner's Dilemma, it would be better for both parties to exit the protracted conflict, but no party has the unilateral incentive to do so. The critical results of this paper do not depend on a prisoner's dilemma structure; they require only that a better result can be achieved through trust or credible commitment. Thus, a stag hunt or other game structures are also possible. The Prisoner's Dilemma is used because of its wide exposure in the literature.\footnote{This payoff matrix does not distinguish between civil and interstate conflict, though in practice the effects on the payoffs might indeed have a different structure in the different cases.}

The countries' stage game actions will be referred to as ``Trust'' and ``Fight'' as represented in Table 1, where Country 1 is the row player and Country 2 is the column player.

\begin{center}
\captionof{table}{Stage game payoffs}
\begin{tabular}{|l|c|c|}
  \hline      & Trust & Fight \\ \hline
	 Trust& $T$, $T$& $-D$, $T+W$ \\ \hline
	Fight & $T+W$, $-D$& $W-D$, $W-D$ \\ \hline
\end{tabular}
\end{center}

Here $T>0$ represents the benefit from a country's negotiating partner playing Trust. $D>0$ represents the damages incurred when a country's negotiating partner plays Fight, while $W>0$ is the additional benefit a country receives when it plays Fight itself. We will refer to the stage-game Nash Equilibrium of (Fight, Fight) as ``War.'' We will refer to (Trust,Trust) as ``Peace.'' In line with the assumption that the countries play a Prisoner's Dilemma in the stage game, we assume $T>W-D$. Note that this ensures that the joint payoffs from ``Peace'' are the highest among all the stage-game action profiles.

@@

We will use grim trigger punishments throughout: that is, (Fight, Fight) will be played forever if Fight is played by any party in any round. Because we use grim trigger punishments, there are only 4 interesting paths of play after concessions are given:
1) Both countries play Trust always; 
2) Both play Fight always; 
3) Country 1 plays Fight in round 1 while country 2 plays Trust, which is followed by both countries playing Fight from round 2 onwards; and 
4) Country 2 plays Fight in round 1 while country 1 plays Trust, which is followed by both countries playing Fight from round 2 onwards.

Equilibrium payoffs beginning from round 1 are now easily described by noting only the collective first round behavior.\footnote{If a parameter of the model changes after period 0, the equilibrium may be disrupted. Although it is beyond the scope of this paper, one could model this by adding a probability of equilibrium disruption followed by a new round of concessions that would essentially re-start the game--albeit with altered parameters.} Let $X_{ij}$ represent the sum of discounted payoffs for country $i$ where subscripts represent first round strategies of that country. If both countries play T in round 1, the corresponding payoff to country $i$ is represented as $ X_{TT}^i=\frac {T}{1- \delta_i}$. Likewise $X_{TF}^i=({-D}+\frac{\delta_i(W-D)}{1- \delta_i} )$, $ X_{FF}^i=\frac{W-D}{1- \delta_i}$, and  $X_{FT}^i=T+W +\frac{\delta_i(W-D)}{1- \delta_i}$. $X_{FT}^h$, for example, represents payoffs for the high type of country 1 when country 1 fights and country 2 trusts in the first period. Note that because types have different discount factors, the payoffs differ by type.

Parameters are common knowledge with the exception of $\delta_i$, which is country $i$'s private information. We will take the measure of social welfare, and thus the determinant of the optimal equilibrium, to be the sum of participating high types' expected utilities.

\section{Analysis of Benchmark Cases}
\label{sec:bench}
When countries are faced with uncertainty about each other's motives, they can proceed in several ways, including giving concessions in order to signal their ability to make peace. We will examine three types of equilibria.\footnote{If the discount factor is high enough, there are many types of equilibria. Consider for instance oscillating every other period between (Trust,Trust) and (Fight,Fight) with a grim trigger ``War'' threat. In this paper, we are not interested in these equilibria because they are not as realistic for applications to conflict scenarios. These are also always payoff dominated by at least one of the other equilibria discussed and thus 	are not attractive.} The Appendix provides a thorough analysis of the various equilibria as well as most proofs.

\subsection{Pooling equilibrium}
In the \emph{pooling equilibrium}, neither negotiating party gives concessions in period 0, then both parties play Fight in every period beginning with period 1. Countries do nothing to bridge the credibility gap, and thus do not cooperate. This \emph{pooling equilibrium} is likely to happen when countries are in long-set patterns of distrust. Active fighting is not necessary; remaining outside of a state of peace is all that is required.  See the Appendix for a formal characterization of this type of equilibrium.

Cyprus' civil war is a good example of the pooling equilibrium where neither side is taking actions to move things forward. While both sides posture that they want peace, and while there has been some lessening of tensions, very few concrete actions have been taken that move the sides toward settlement. This conflict is a stalemate.

>>>>

\begin{abstract}
When two parties are engaged in conflict and each distrusts the other's ability to cooperate or make peace, concessions can be used to indicate an interest in cooperation or peacemaking. However, when negotiating parties are concerned that concessions could be used against them in the future, a lack of trust can prevent optimal concessions from being made. It can also reduce the possibility of peace or cooperation. Using a repeated game that is preceded by an opportunity to signal one's commitment to cooperation through the provision of concessions, we formally demonstrate that concerns over the future use of concessions can explain the existence of inefficient concessions. We then use mechanism design to explore ways in which a third party mediator can act as a guarantor that promised concessions will be delivered, thereby reducing inefficiencies and increasing the potential for peace and cooperation. In this process, we open up a new rationale for mediation: to increase the efficiency of signaling in a preliminary round of negotiations and to overcome the concern that concessions will be used against the giver in the future. 
\end{abstract}

@@

In negotiations with its neighbors, Israel is hesitant to give certain concessions. The Golan Heights, captured from Syria during the 1967 War, is a particular point of contention in Israeli-Syrian relations. While the Golan Heights does not have the West Bank’s historic or religious importance, Israel values it for its highly strategic location. The Golan Heights could be a useful concession, but Israel senses that were she to give over the land, Syria may ultimately use the concession against her. This same hesitancy exists with the West Bank settlements. Many are located on hilltops that could offer strategic military outposts for forces hostile to Israel. When she occasionally abandons settlements, Israel generally dismantles all infrastructure, including water and electricity, and bulldozes the buildings. Although inefficient, this reduces the possibility of those settlements being used against Israeli interests. Without trust between negotiating parties, the most efficient concessions often are often difficult to arrange.

In both active wars and long-simmering conflicts, settlement negotiations often occur simultaneously with continued fighting. This is true in non-violent disputes as well, such as long-running episodes of geo-strategic competition: parties engage in costly conflict while they negotiate, foregoing the value of the public goods they could create through cooperation. If settlement can be achieved, additional costs of conflict -- including additional death and destruction -- are averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (e.g., Bercovitch 2000; Greig 2005a; Beardsley et al. 2006; Duursma 2020).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that conflict participants can offer. In our context, the efficiency of a concession is determined by how much of the concessions's material value is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

We introduce a two-sided signaling model followed by a repeated Prisoner's Dilemma in which concessions can provide future value to the recipient and/or serve as a signal that provides the recipient with information about the type of player the conceding state is. We use this model to show that concerns surrounding the possibility that concessions could be used against the giver in the future can undermine cooperation. Using a mechanism design framework, we demonstrate that a mediator can facilitate more efficient concessions by removing uncertainty about the ability of the parties to commit to a settlement. Perhaps more importantly, the mediator can also sometimes facilitate settlement when the negotiating parties cannot achieve it through bilateral interaction. In other words, commitment is essentially a far-sighted preference that allows cooperation to emerge, uncertainty can result in missed opportunities for cooperation, and concessions and mediation can facilitate conflict settlement and the provision of public goods.
%The sentence above literally parrots the editor's words back to them. Too on the nose?

%Attempting to add in the China-US case in a prominent way. The prose is more awkward than I would like still. - March 13, 2023 BG
Three archetypal situations illustrate the significance and difficulty of efficient concession-making in dispute settlement.  The first is the case of disarmament, the second is the concession of strategically valuable territory, and the third is the export of advanced technology to a strategic rival. %In each case, efficient concessions are difficult to make because the item conceded may be used by the recipient to bolster their military capabilities, altering the balance of power in the ongoing conflict. Also true in each case, if dispute settlement is achieved, a public good is produced -- peace in the first two cases and innovation and growth in the civilian economy in the third case.

Barbara Walter identified disarmament as the ``critical barrier to civil war settlement'' (1997). Particularly in conflicts between a rebel group and a central government, the best-achievable negotiated settlement often involves policy concessions by the government in exchange for disarmament and recognition of government authority by the rebels. Disarmament by rebels is an efficient concession if fully-operational weapons are turned over to the central government for its own use, while disarmament accompanied by decommissioning is inefficient because the value of the arms is destroyed. Disarming and turning over fully functional weapons is an incredibly difficult concession for rebels to make: once they turn over their arms, they lose future bargaining leverage and have no guarantee that their own weapons will not be used against them later. If the rebels can instead decommission those weapons, an important stumbling block to peace is removed.\footnote{Other stumbling blocks remain, including the fact that the government retains its weapons. Thus, credible commitment on the part of the government remains difficult, even if decommissioning occurs.} Peace may therefore be possible with inefficient concessions where it is impossible with efficient concessions.\footnote{Walter's approach is part of a much broader literature on the bargaining model of war that looks at concessions and asymmetric information, including  Fearon (1996), Powell (1996a and 1996b, 2006), Slantchev (2011) and Arena (2013).}

Control of strategically valuable territory can pose similar challenges in both civil and interstate wars. Even if transferring control of strategically valuable territory creates the best chance for peace, its transfer often involves a substantial risk that the territory will provide military advantage in a future conflict. Following the 1992 war between Armenia and Azerbaijan over the disputed region of Ngorno-Karabakh, Armenia retained control of a significant amount of Azeribaijani territory that neither the Armenian government nor the Karabakh government viewed to be part of its homeland. Azerbaijan, however, valued control of the territory greatly as the home of Azeri residents and wished to govern the territory. Thus, it might have been efficient in the mid-1990s for the Armenian/Karabahk side to cede control of that territory to Azerbaijan as part of a deal in which Azerbaijan recognized Karabakh's independence.  However, that territory was strategically valuable, including territory on both sides of the Lachin Corridor, which connects Ngorno-Karabakh and Armenia. Azerbaijani control of the territory surrounding the corridor would empower Azerbaijan to seize military advantage in the future and cut off land travel between Ngorno-Karabakh and Armenia.  The territory was not conceded and the conflict persisted in costly stalemate for almost twenty years until war broke out anew in 2020.

In our third archetypal example, under both the Trump and Biden administrations, the US has restricted the export of cutting-edge technology to China. China and the US are engaged in geo-strategic competition. Both countries could benefit from the more rapid innovation and growth that free trade in chips and other technology would enable, but there is risk that any technology exported will be used in the military sector to pursue advanced weapons, altering the military balance of power. As of 2023, trade in numerous advanced technologies is restricted, reducing the global pace of innovation and economic growth.

In all three of these archetypal examples, cooperative/peaceful outcomes are extremely challenging to reach. Walter (1997) observes empirically that, due to the difficulty of making efficient disarmament concessions, civil wars tend to end on the battlefield, rather than through negotiated settlement. Interstate wars over territory are similarly prone to violent, rather than negotiated, resolution (Hensel 1996), and the trajectory of competition and cooperation between the US and China remains highly uncertain. Can mediation offer a path to more settlement, peace, and prosperity in these challenging contexts?

This paper provides an explanation for why efficient concessions are sometimes impossible to make. It also revises our understanding of why inefficient concessions can sometimes be beneficial and how mediators can both help increase the efficiency of concessions and make cooperation possible. Weapons, strategically valuable territory, and the sharing of advanced technology will not be conceded if the conceding side fears that these items may be used in future attacks against them. Countries are reluctant to make concessions that can be used against them in latter rounds of conflict or negotiation. One way to mitigate the risk of dangerous concessions is to destroy the future value of the concession---e.g. decommission weapons, attempt to reduce the strategic value of the territory being conceded, or only allow trade in less advanced technologies or technologies without military applications. Another option, and the one on which we focus in this paper, is third-party mediation. We use this model to demonstrate that a third-party mediator can facilitate negotiated settlement by increasing trust and enabling disputants to make more efficient concessions. The mediator does this by only requiring concessions of negotiating parties who are able to cooperate, thereby removing the risk that concessions might be used against the giver in the future.

%While an integral part of the literature on conflict, concessions have been only partially addressed by theoretical research. The broader literature that attempts to model war has included concessions as part of bargaining, but has not explicitly considered the role of concessions in achieving peace. Schelling's \emph{Arms and Influence} (1966) modeled war with a game theoretic framework, and Powell (1999) has also defended the use of modeling in understanding war. Fearon (1996), Powell (1996a and 1996b, 2006) and others consider war to be a bargaining failure, caused by commitment problems and asymmetric information. When the assets that are the object of the bargaining confer future power, a lack of credible commitment, or trust, can create a holdup problem that prevents the optimal level of cooperation (Fearon 1996). %[OUTSIDE THE WAR CONTEXT..., ???]

%In these models war can be averted by splitting the pie (e.g., the disputed territory) in a way acceptable to both parties. Concessions are modeled as offering part of the pie to the other party, and asymmetric information is about each party's \textit{resolve} to fight and/or military capabilities (Arena 2013). In this context, an example of costly signaling would be military preparations for war (Slantchev 2011).

%We approach the commitment problem through this same general understanding, but

We model a party's ability to cooperate as private information. A high type is able to cooperate; a low type has incentives not to cooperate. These types map onto the commitment problem, with a high type one who is able to commit. A low type, though perhaps willing, is unable to do so. A low type may posture like the high type, but lacks the political muscle or will to follow through on a peace agreement. Democratic leaders may more often be high types because they are less likely to face coups or other hostilities in reaction to short term decisions. This is not always the case, however. Mahmoud Abbas, President of the Palestinian National Authority, may be a low type with respect to his negotiations with Israel, not because he does not want to make peace, but because he cannot credibly act on behalf of the entire Palestinian population.

In our model, concessions may act as signals or as an end in themselves -- or as both simultaneously. One contribution of this paper is that it allows concessions to have future material value, signaling value, and both. We will show that when concessions have both signaling and future material value, some of the material value of the concessions may be purposefully destroyed and thus inefficiencies can emerge. Another contribution is that the way we model concessions does not require countries to make an agreement that requires credible commitment, as is common in bargaining models. Instead, countries willingly provide a concession if they expect the signal it sends to provide a benefit in terms of future cooperation and public good provision.\footnote{When mediation is required to achieve cooperation in the model, it relies upon a manipulative mediator to enforce the required level of concessions.}

In analyzing both the signaling and material value of concessions, an extensive literature on gift giving is particularly helpful. Like concessions, costly gifts have been shown to have a valuable role in relationship building. Furthermore, the gift giving literature---more so than the concessions literature---focuses specifically on the issue of efficiency. Inefficiency is defined for the purposes of this paper as giving a gift (i.e. a concession) that is more costly to the giver than it is valuable to the recipient.

%I need to add a bit about historical international relations, and a couple of cites, to the two paragraphs below. BG 3-13-2023
Often, inefficient gifts occur in the formative stages of relationships or in immature relationships (Camerer 1988). Gifts given in established relationships, such as wedding presents or spousal gifts, are more likely to be cash gifts or efficient gifts that are specifically requested or selected from a registry. Since inefficient gift giving is more likely involved in nascent relationships and partnerships, it is natural to suppose that inefficient gifts play a signaling role in addition to the material value of transferring the gift's inherent worth.

In efforts to explain the existence of gift giving,  Camerer (1988) and Van de Ven (2000) give several anthropological explanations for inefficiency.\footnote{Prendergast and Stole (2001) also provide an explanation for inefficient gift giving focused on the utility of matching.} Amongst these explanations are altruism, social mores, and egoism. But the most relevant explanation for the field of international relations is that gift giving is strategic. Camerer uses mechanism design to model gifts as costly signals, where inefficiency is useful in pre-play communication in order to form relationships between like types. This strategic giving of gifts in societies looks very much like the strategic giving of concessions to remedy conflict situations. When modeled game-theoretically, these environments look even more similar.

@@

Camerer shows that in this general formulation, a costly signaling model will always yield efficient gifts.\footnote{In an environment with repeated opportunities for concessions, Watson (1999) is able to show that a strong relationship can be reached by starting with small ``gifts'' and increasing their size as the relationship becomes more committed. See Van de Ven (2000) for an overview of this literature.} In order to explain inefficiency, Camerer adds an additional pre-play period where players must pay a cost in order to enter the game and send and receive gifts. If types separate in the pre-play period with low types not being willing to pay, high types are saved the cost of sending a signal to low types in the main game.  High types would, in some parameterizations, be given incentives to give inefficient equilibrium gifts in the main game in order to lower the low types' expected payoff of playing to below the pay-to-play fee. Such an action can keep out the low types and this can raise the expected payoff for a high type by saving on the gifts given.

%Prendergast and Stole (2001) provide a second explanation for inefficient gifts. The authors explain inefficient gift giving by modeling the utility of matching. That is, if players are of a like mindset they get an extra benefit so they are willing to take the risk of giving the wrong gift.

We propose a different conceptual framework to explain this real-life phenomenon that is rooted in a realistic international relations puzzle. The basic formulation of our mechanism-design approach is an expansion of Camerer's (1988) analysis. The model can achieve the basic `inefficient gifts' results of Camerer (1988) %and Prendergast and Stole (2001) 
but does so using a very different explanation for inefficiency. The basic model is a costly signaling model, at its root similar to Spence (1973) and subsequent papers.\footnote{See Connelly et al. (2011) for a review of the costly signaling literature.} In essence, the model shows that if concessions can be later used against the party that gives them, then it can be in a negotiating party's best interest to give inefficient concessions. While our model is tailored to an international relations setting, the explanation for inefficient giving of concessions may be more broadly applicable.

@@

It is essential to focus more precisely on the specific actions and environments that make use of particular types of mediation. To this end, this paper models an environment where the primary issue is trust, not uncertainty about an opponent's capabilities, and where a mediator is employed who has manipulative abilities. Parties are more likely to seek mediation when they are entrenched in costly conflict (Bercovitch and Jackson 2001; Greig 2005b; Greig and Diehl 2006; Terris and Maoz 2005; Svensson 2008); we thus model parties to a conflict who can benefit from seeking mediation in order to overcome mistrust.\footnote{Alternatively, parties may seek a mediator to justify difficult concessions and avoid angering domestic constituencies (Allee and Huth 2006, Beardsley 2010; Beardsley and Lo 2014).}

These empirical results have until recently lacked strong theoretical foundations. Mediation is not uncommon, but we are only beginning to understand why actors use it. Recently, mediation has begun to be modeled using mechanism design for settings where militarization or the resolve to fight are private information (Bester and Warneryd 2006; Fey and Ramsay 2009, 2010 and 2011; Horner et. al. 2010; Meirowitz et al. 2012, 2019). In particular, Meirowitz et al. 2019 shows that---when asymmetric information is about the level of militarization---a `Myerson' mediator can circumvent the incentives for militarization that arise in unmediated peace talks and improve the changes for peace. This paper expands the theoretical literature by modeling the role of a manipulative mediator when the information asymmetry is about parties' abilities to commit to cooperation and concessions have future material value that could be used to harm the giver. Here, a mediator can achieve peace or the provision of other public goods (1) in situations where bilateral concessions cannot, and (2) with concessions that are more efficient when bilateral concessions must be inefficient to achieve peace.

@@

Beginning in period 1, countries engage in an infinitely repeated Prisoner's Dilemma once per period with stage game payoffs in Table 1 below.\footnote{For a game-theoretic analysis of the repeated Prisoner's Dilemma with uncertainty over discount factors, focused on studying players' belief structures, see Maor and Solan (2015).  Our model differs in that it adds an initial period, before the repeated Prisoner's Dilemma begins, in which players make costly concessions to each other.  We will see that these concessions enable signaling between the players that render the analysis of their beliefs much simpler than in Maor and Solan (2015).} While not a good representation of all-out war or even a zero-sum short term war, the Prisoner's Dilemma is appropriate to represent a limited, continuing conflict--military or otherwise. In cases appropriately modeled by a Prisoner's Dilemma, it would be better for both parties to exit the protracted conflict, but no party has the unilateral incentive to do so. The critical results of this paper do not depend on a prisoner's dilemma structure; they require only that a better result can be achieved through trust or credible commitment. Thus, a stag hunt or other game structures are also possible. The Prisoner's Dilemma is used because of its wide exposure in the literature.\footnote{This payoff matrix does not distinguish between civil and interstate conflict, though in practice the effects on the payoffs might indeed have a different structure in the different cases.}

The countries' stage game actions will be referred to as ``Trust'' and ``Distrust'' as represented in Table 1, where Country 1 is the row player and Country 2 is the column player.

\begin{center}
\captionof{table}{Stage game payoffs}
\begin{tabular}{|l|c|c|}
  \hline      & Trust & Distrust \\ \hline
	 Trust& $T$, $T$& $-D$, $T+W$ \\ \hline
	Distrust & $T+W$, $-D$& $W-D$, $W-D$ \\ \hline
\end{tabular}
\end{center}

Here $T>0$ represents the benefit from a country's negotiating partner playing Trust. $D>0$ represents the damages incurred when a country's negotiating partner plays Distrust, while $W>0$ is the additional benefit a country receives when it plays Distrust itself. We will refer to the stage-game Nash Equilibrium of (Distrust, Distrust) as No Cooperation. We will refer to (Trust,Trust) as Cooperation. In line with the assumption that the countries play a Prisoner's Dilemma in the stage game, we assume $T>W-D$. Note that this ensures that the joint payoffs from Cooperation are the highest among all the stage-game action profiles.

@@

We will use grim trigger punishments throughout: that is, (Distrust, Distrust) will be played forever if Distrust is played by any party in any round. Because we use grim trigger punishments, there are only 4 interesting paths of play after concessions are given:
1) Both countries play Trust always; 
2) Both play Distrust always; 
3) Country 1 plays Distrust in round 1 while country 2 plays Trust, which is followed by both countries playing Distrust from round 2 onwards; and 
4) Country 2 plays Distrust in round 1 while country 1 plays Trust, which is followed by both countries playing Distrust from round 2 onwards.

Equilibrium payoffs beginning from round 1 are now easily described by noting only the collective first round behavior.\footnote{If a parameter of the model changes after period 0, the equilibrium may be disrupted. Although it is beyond the scope of this paper, one could model this by adding a probability of equilibrium disruption followed by a new round of concessions that would essentially re-start the game--albeit with altered parameters.} Let $X_{ij}$ represent the sum of discounted payoffs for country $i$ where subscripts represent first round strategies of that country. If both countries play T in round 1, the corresponding payoff to country $i$ is represented as $ X_{TT}^i=\frac {T}{1- \delta_i}$. Likewise $X_{TF}^i=({-D}+\frac{\delta_i(W-D)}{1- \delta_i} )$, $ X_{FF}^i=\frac{W-D}{1- \delta_i}$, and  $X_{FT}^i=T+W +\frac{\delta_i(W-D)}{1- \delta_i}$. $X_{FT}^h$, for example, represents payoffs for the high type of country 1 when country 1 plays Distrust and country 2 plays Trust in the first period. Note that because types have different discount factors, the payoffs differ by type.

Parameters are common knowledge with the exception of $\delta_i$, which is country $i$'s private information. We take achieving peace, or more broadly, achieving cooperation with the provision of public goods, as the normative goal with respect to advancing social welfare. Only high types are capable of making peace/providing public goods, and thus we will take the measure of social welfare to be the sum of the participating high types’ utilities.

\section{Analysis of Benchmark Cases}
\label{sec:bench}
When countries are faced with uncertainty about each other's motives, they can proceed in several ways, including giving concessions in order to signal their ability to Cooperate. We will examine three types of equilibria.\footnote{If the discount factor is high enough, there are many types of equilibria. Consider for instance oscillating every other period between (Trust,Trust) and (Distrust,Distrust) with a grim trigger Distrust threat. In this paper, we are not interested in these equilibria because they are not as realistic for applications to conflict scenarios. These are also always payoff dominated by at least one of the other equilibria discussed and thus 	are not attractive.} The Appendix provides a thorough analysis of the various equilibria as well as most proofs.

\subsection{Pooling equilibrium}
In the \emph{pooling equilibrium}, neither negotiating party gives concessions in period 0, then both parties play Distrust in every period beginning with period 1. Countries do nothing to bridge the credibility gap, and thus do not cooperate. This \emph{pooling equilibrium} is likely to happen when countries are in long-set patterns of distrust. Active fighting is not necessary; remaining outside of a state of peace/cooperation is all that is required.  See the Appendix for a formal characterization of this type of equilibrium.
